---
title: "Data 4010 Test Segmentation"
author: "Patrick Desmarais"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Importing library
library("lidR")
library("rgl")
library("data.table")
```

```{r}
# Importing cropped data 
import <- readLAS("C:/4010/F_240717_1_RGBPC_cropped.las")
treeDataset <- import@data
```

First we will do a regular K-means clustering which will also some what do segmentation.
This first one will include the X and Y

```{r}
# Remove all the columns exept for the coordinates X, Y, and Z and the colours.
cleanDataX <- treeDataset[, c("X", "Y", "Z", "R", "G", "B")]

# We will have to normalize the data before inputting into the K-Means algorithm
# We will use a min-max function
minMax <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}

normDataX <- as.data.frame(lapply(cleanDataX, minMax))

# Now we create the k-means algorithm to split the trees, building, roads and grass
kmDataX <- kmeans(normDataX, centers = 4)


kmDataX$betweenss
kmDataX$withinss

```

```{r}
# Create a copy of the imported data
copyX <- data.frame(treeDataset)

# Merge the K-means cluster and the tree data
newCopyX <- cbind(copyX, kmDataX$cluster)

newCopyX <- setDT(newCopyX)

import@data <- newCopyX

plot(import, color = "kmDataX$cluster")
```

We can see that the k-means segmentation properly separated the buildings and road 
from the trees however we split the trees into two different sections. We can also see
that the algorithm had a hard time differentiating between grass and trees since we put the 
importance of the X and Y values on the same level as the Z axis

Next we will do another k-means however we will remove the X and Y coordinates
```{r}
cleanDataZ <- treeDataset[, c("Z", "R", "G", "B")]

normDataZ <- as.data.frame(lapply(cleanDataZ, minMax))

# Now we create the k-means algorithm to split the trees, building, roads and grass
kmDataZ <- kmeans(normDataZ, centers = 4)


kmDataZ$betweenss
kmDataZ$withinss
```

```{r}
# Create a copy of the imported data
copyZ <- data.frame(treeDataset)

# Merge the K-means cluster and the tree data
newCopyZ <- cbind(copyZ, kmDataZ$cluster)

newCopyZ <- setDT(newCopyZ)

import@data <- newCopyZ

plot(import, color = "kmDataZ$cluster")
```

After removing the X and Y values from the K-means algorithm we can see an improvement
in differentiating between the grass and the trees. We can also see the canopy of the trees

